---
title: "Notes: Introduction to Nonparametric Bootstrapping"
author: "Ivan Jacob Agaloos Pesigan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes: Introduction to Nonparametric Bootstrapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(jeksterslabRboot)
```

## Random Variable

```{r, echo = FALSE}
set.seed(42)
n <- 1000
df <- n - 1
B <- 1000
mu <- 100
theta <- mu
sigma2 <- 225
sigma <- sqrt(sigma2)
var_thetahat <- sigma2 / n
se_thetahat <- sqrt(var_thetahat)
```

Let $X$ be a random variable
from a normal distribution with $\mu = `r mu`$
and $\sigma^2 = `r sigma2`$.

\begin{equation}
  X
  \sim
  \mathcal{N}
  \left(
    \mu = `r mu`,
    \sigma^2 = `r sigma2`
  \right)
\end{equation}

This data generating function is referred to as $F$.

```{r, echo = FALSE}
Variable <- c(
  "`theta`",
  "`sigma2`",
  "`sigma2`"
)
Description <- c(
  "Population mean.",
  "Population variance.",
  "Population standard deviation."
)
Notation <- c(
  "$\\theta = \\mu$",
  "$\\sigma^2$",
  "$\\sigma$"
)
Value <- c(
  theta,
  sigma2,
  sigma
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Population Parameters"
)
```

For a sample size of
$n = `r n`$,
using the know parameters,
the variance of the sampling distribution of the mean is

\begin{equation}
  \mathrm{Var}
  \left(
    \hat{\mu}
  \right)
  =
  \frac{
    \sigma^2
  }
  {
    n
  }
  =
  `r var_thetahat`
\end{equation}

\noindent and the standard error of the mean is

\begin{equation}
  \mathrm{se}
  \left(
    \hat{\mu}
  \right)
  =
  \frac{
    \sigma
  }
  {
    \sqrt{n}
  }
  =
  `r se_thetahat`
\end{equation}

```{r, echo = FALSE}
Variable <- c(
  "`n`",
  "`theta`",
  "`var_thetahat`",
  "`se_thetahat`"
)
Description <- c(
  "Sample size.",
  "Population mean.",
  "Variance of the sampling distribution of the mean.",
  "Standard error of the mean."
)
Notation <- c(
  "$n$",
  "$\\theta = \\mu$",
  "$\\mathrm{Var} \\left( \\hat{\\theta} \\right) = \\frac{ \\sigma^2 }{n}$",
  "$\\mathrm{se} \\left( \\hat{\\theta} \\right) = \\frac{ \\sigma }{\\sqrt{n}}$"
)
Value <- c(
  n,
  theta,
  var_thetahat,
  se_thetahat
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Sampling Distribution of $\\hat{\\theta}$ with Known Parameters"
)
```

## Realizations of the Random Variable

Let $\mathbf{x}$ be a vector of length $n = `r n`$
of realizations of the random variable $X$.
With the
`rnorm()`
function
we generate a vector
`x`
with
`n =` `r n`,
`mean =` `r mu`,
and
`sd =` `r sigma`.

```{r}
x <- rnorm(
  n = n,
  mean = mu,
  sd = sigma
)
```

We can calculate sample statistics using the sample data $\mathbf{x}$.
Note that a hat (`^`) indicates that the quantity is an estimate of the parameter
using sample data.

```{r, echo = FALSE}
muhat <- mean(x)
thetahat <- muhat
sigma2hat <- var(x)
sigmahat <- sqrt(sigma2hat)
varhat_thetahat <- sigma2hat / n
sehat_thetahat <- sqrt(varhat_thetahat)
Variable <- c(
  "`n`",
  "`thetahat`",
  "`sigma2hat`",
  "`sigmahat`",
  "`varhat_thetahat`",
  "`sehat_thetahat`"
)
Description <- c(
  "Sample size.",
  "Sample mean.",
  "Sample variance.",
  "Sample standard deviation.",
  "Estimate of the variance of the sampling distribution of the mean.",
  "Estimate of the standard error of the mean."
)
Notation <- c(
  "$n$",
  "$\\hat{\\theta} = \\hat{\\mu} = \\frac{1}{n} \\sum_{i = 1}^{n} x_i$",
  "$\\hat{\\sigma}^2 = \\frac{1}{n - 1} \\sum_{i = 1}^{n} \\left( x_i - \\hat{\\mu} \\right)^2$",
  "$\\hat{\\sigma} = \\sqrt{\\frac{1}{n - 1} \\sum_{i = 1}^{n} \\left( x_i - \\hat{\\mu} \\right)^2}$",
  "$\\hat{\\mathrm{Var}} \\left( \\hat{\\theta} \\right) = \\frac{ \\hat{\\sigma}^2 }{n}$",
  "$\\hat{\\mathrm{se}} \\left( \\hat{\\theta} \\right) = \\frac{ \\hat{\\sigma} }{\\sqrt{n}}$"
)
Value <- c(
  n,
  thetahat,
  sigma2hat,
  sigmahat,
  varhat_thetahat,
  sehat_thetahat
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Sample Statistics (Parameter Estimates)"
)
```

```{r, echo = FALSE}
hist(x)
qqnorm(x)
qqline(x)
```

## Nonparametric Bootstrapping

$\mathbf{x}$ which is a vector of realizations of the random variable $X$
is used in bootstrapping.
This sample data is referred to as the empirical distribution
$\hat{F}$.

In nonparametric bootstrapping,
the empirical distribution
$\hat{F}$
is resampled
$B$ number of times (`r B` in the current example) with replacement.

### $x^{*} = \left\{ x^{*1}, x^{*2}, x^{*3}, x^{*b}, \dots, x^{*B} \right\}$

$x^{*}$ is a set of $B$ bootstrap samples.

The function `nb()`
from the `jeksterslabRboot` package
can be used to generate `B`
nonparametric bootstrap samples.
The argument `data` takes in a vector, matrix, or data frame.
`data` is the empirical distribution
$\hat{F}$.
In the example below,
we use `x`, which is the sample data above,
as the argument `data`.
The argument `B` takes an integer.
`B` is the number of bootstrap samples
`B =` `r B` in our case.
Optional arguments include
`par` which is a logical argument to enable the use
of multiple cores
and `ncores` which is the number of cores to use.
The `nb()` function returns a `list`
of length `B` bootstrap samples.
$x^{*}$ is saved as `x_star`.

```{r}
x_star <- nb(
  data = x,
  B = B
)
```

### $\hat{\theta}^{*} = \left\{ \hat{\theta}^{*} \left( 1 \right), \hat{\theta}^{*} \left( 2 \right), \hat{\theta}^{*} \left( 3 \right), \hat{\theta}^{*} \left( b \right), \dots, \hat{\theta}^{*} \left( B \right)\right\}$

The parameter is estimated for each of the $B$
bootstrap samples.
For example,
if we are interested in making inferences about the mean,
we estimate the mean $B$ times.
The estimated parameters form the bootstrap sampling distribution
$\hat{\theta}^{*}$.

`sapply` is a useful tool in `R`
to apply a function, in our case `mean`,
to each element of a list,
in our case `x_star`.
`sapply` simplifies the output when possible.
In our case,
`sapply` returns a vector,
`thetahat_star`.
The `apply` family of functions
as well as their parallel counterparts
(e.g., `parSapply`)
can be used to simplify repetitive tasks.

```{r}
thetahat_star <- sapply(
  X = x_star,
  FUN = mean
)
```

```{r, echo = FALSE}
hist(
  thetahat_star,
  main = expression(
    paste(
      "Histogram of ",
      hat(theta),
      "*"
    )
  ),
  xlab = expression(
    paste(
      hat(theta),
      "*"
    )
  )
)
qqnorm(thetahat_star)
qqline(thetahat_star)
```

### $\hat{\mathrm{se}}_{\mathrm{B}} \left( \hat{\theta} \right)$

```{r, echo = FALSE}
mean_thetahat_star <- mean(thetahat_star)
var_thetahat_star <- var(thetahat_star)
sd_thetahat_star <- sqrt(var_thetahat_star)
```

The estimated bootstrap standard error
is given by

\begin{equation}
  \hat{\mathrm{se}}_{\mathrm{B}}
  \left(
    \hat{\theta}
  \right)
  =
  \sqrt{
    \frac{1}{B - 1}
    \sum_{b = 1}^{B}
    \left[
      \hat{\theta}^{*} \left( b \right)
      -
      \hat{\theta}^{*} \left( \cdot \right)
    \right]^2
  }
  =
  `r sd_thetahat_star`
\end{equation}

\noindent where

\begin{equation}
  \hat{\theta}^{*}
  \left(
    \cdot
  \right)
  =
  \frac{1}{B}
  \sum_{b = 1}^{B}
  \hat{\theta}^{*} \left( b \right) 
  =
  `r mean_thetahat_star`
  .
\end{equation}

Note that
$\hat{\mathrm{se}}_{\mathrm{B}} \left( \hat{\theta} \right)$
is the standard deviation of
$\hat{\theta}^{*}$
and
$\hat{\theta}^{*} \left( \cdot \right)$
is the mean of
$\hat{\theta}^{*}$ .

```{r, echo = FALSE}
Variable <- c(
  "`B`",
  "`mean_thetahat_star`",
  "`var_thetahat_star`",
  "`sd_thetahat_star`"
)
Description <- c(
  "Number of bootstrap samples.",
  "Mean of $B$ sample means.",
  "Variance of $B$ sample means.",
  "Standard deviation of $B$ sample means."
)
Notation <- c(
  "$B$",
  "$\\hat{\\theta}^{*} \\left( \\cdot \\right) = \\frac{1}{B} \\sum_{b = 1}^{B} \\hat{\\theta}^{*} \\left( b \\right)$",
  "$\\hat{\\mathrm{Var}}_{\\mathrm{B}} \\left( \\hat{\\theta} \\right) = \\frac{1}{B - 1} \\sum_{b = 1}^{B} \\left[ \\hat{\\theta}^{*} \\left( b \\right) - \\hat{\\theta}^{*} \\left( \\cdot \\right) \\right]^2$",
  "$\\hat{\\mathrm{se}}_{\\mathrm{B}} \\left( \\hat{\\theta} \\right) = \\sqrt{ \\frac{1}{B - 1} \\sum_{b = 1}^{B} \\left[ \\hat{\\theta}^{*} \\left( b \\right) - \\hat{\\theta}^{*} \\left( \\cdot \\right) \\right]^2 }$"
)
Value <- c(
  B,
  mean_thetahat_star,
  var_thetahat_star,
  sd_thetahat_star
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Nonparametric Bootstrapping Results"
)
```

### Confidence Intervals

Confidence intervals can be constructed
around
$\hat{\theta}$ .
The functions
`pc()`,
`bc()`,
and
`bca()`
from
the `jeksterslabRboot` package
can be used to construct
confidence intervals
using the default alphas
of
$0.001$,
$0.01$,
and
$0.05$.
Wald confidence intervals (`wald()`)
are added for comparison.
The confidence intervals can also be evaluated.
See
`zero_hit()`,
`theta_hit()`,
`len()`,
and
`shape()`
from
the `jeksterslabRboot` package.
Since we know the population parameter theta
$\left(\theta = \mu = `r theta` \right)$,
we can check if our confidence intervals contain
the population parameter.

```{r}
wald_out <- wald(
  thetahat = thetahat,
  sehat_thetahat = sehat_thetahat,
  eval = TRUE,
  theta = theta
)
wald_out_t <- wald(
  thetahat = thetahat,
  sehat_thetahat = sehat_thetahat,
  distribution = "t",
  df = df,
  eval = TRUE,
  theta = theta
)
pc_out <- pc(
  thetahat_star = thetahat_star,
  thetahat = thetahat,
  wald = TRUE,
  eval = TRUE,
  theta = theta
)
bc_out <- bc(
  thetahat_star = thetahat_star,
  thetahat = thetahat,
  wald = TRUE,
  eval = TRUE,
  theta = theta
)
bca_out <- bca(
  thetahat_star = thetahat_star,
  thetahat = thetahat,
  data = x,
  fitFUN = mean,
  wald = TRUE,
  eval = TRUE,
  theta = theta
)
```

```{r}
knitr::kable(
  x = as.data.frame(
    rbind(
      wald = wald_out,
      wald_t = wald_out_t,
      pc = pc_out,
      bc = bc_out,
      bca = bca_out
    )
  ),
  caption = "Confidence Intervals"
)
```

## Notes

The estimated bootstrap standard error
is similar to results from the closed form solution.
The estimated bootstrap confidence intervals
are also close to the Wald confidence intervals.
While a closed form solution for the standard error is available
in this simple example,
bootstrap standard errors and confidence intervals
can be used in situations
when standard errors are not easy to calculate.
