---
title: "Notes: Nonparametric Bootstrapping - Mediation"
author: "Ivan Jacob Agaloos Pesigan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes: Nonparametric Bootstrapping - Mediation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

THIS IS AN INITIAL DRAFT.


```{r}
library(jeksterslabRboot)
library(MASS)
```

## Random Variables

### Population Model

```{r, echo = FALSE}
n <- 1000
B <- 5000
varnames <- c("X", "M", "Y")
tauprime <- 0.10
beta <- 0.10
alpha <- 0.70
alphabeta <- alpha * beta
theta <- alphabeta
sigma2X <- 225
sigma2M <- 225
sigma2Y <- 225
sigma2epsilonM <- sigma2M - (alpha^2 * sigma2X)
sigma2epsilonY <- sigma2Y - ((beta^2 * alpha^2 * sigma2X) + (beta^2 * sigma2epsilonM) + (2 * alpha * beta * tauprime * sigma2X) + (tauprime^2 * sigma2X))
muX <- 100
muM <- 100
muY <- 100
deltaM <- muM - (alpha * muX)
deltaY <- muY - ((tauprime * muX) + (beta * deltaM) + (alpha * beta * muX))
A <- matrix(
  data = c(
    0,
    alpha,
    tauprime,
    0,
    0,
    beta,
    0,
    0,
    0
  ),
  ncol = 3
)
colnames(A) <- varnames
rownames(A) <- varnames
S <- matrix(
  data = c(
    sigma2X,
    0,
    0,
    0,
    sigma2epsilonM,
    0,
    0,
    0,
    sigma2epsilonY
  ),
  ncol = 3
)
colnames(S) <- varnames
rownames(S) <- varnames
F <- diag(3)
colnames(F) <- varnames
rownames(F) <- varnames
I <- diag(3)
colnames(I) <- varnames
rownames(I) <- varnames
Sigmatheta <- F %*% solve(I - A) %*% S %*% t(solve(I - A)) %*% t(F)
colnames(Sigmatheta) <- varnames
rownames(Sigmatheta) <- varnames
mutheta <- c(
  muX,
  muM,
  muY
)
names(mutheta) <- varnames
```

\begin{equation}
  Y
  =
  \delta_Y
  +
  \tau^{\prime}
  X
  +
  \beta
  M
  +
  \varepsilon_{Y}
\end{equation}

\begin{equation}
  M
  =
  \delta_M
  +
  \alpha
  X
  +
  \varepsilon_{M}
\end{equation}

```{tikz, simple_mediation, echo = FALSE, fig.cap = "The Simple Mediation Model", fig.ext = "png", cache = TRUE}
\usetikzlibrary{er, arrows, positioning}
\begin{tikzpicture}[
  auto,
  node distance = 20mm,
  manifest/.style = {
    rectangle,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  inv/.style = {
    rectangle,
    draw=none,
    fill=none,
    inner sep = 0pt,
    minimum width = 15mm,
    minimum height = 10mm
  },
  error/.style = {
    ellipse,
    draw,
    thick,
    inner sep = 0pt,
    minimum size = 7mm,
    align = center
  },
  mean/.style={
    regular polygon,
    regular polygon sides = 3,
    draw,
    thick,
    inner sep = 0pt,
    minimum width = 7mm,
    minimum height = 7mm
  },
  path/.style = {
    ->,
    thick,
    >=stealth'
  },
  cov/.style = {
    <->,
    thick,
    >=stealth'
  },
]
  \node[manifest] (X) {$X$};
  \node[mean] (1) [above right = of X] {$1$};
  \node[manifest] (M) [above = of 1] {$M$};
  \node[manifest] (Y) [below right = of 1] {$Y$};
  \node[error] (epsilon_M) [right = of M] {$\epsilon_M$};
  \node[error] (epsilon_Y) [right = of Y] {$\epsilon_Y$};
  \draw [path] (X) to node {$\tau^{\prime}$} (Y);
  \draw [path] (X) to node {$\alpha$} (M);
  \draw [path] (M) to node {$\beta$} (Y);
  \draw [path] (epsilon_M) to node {$1$} (M);
  \draw [path] (epsilon_Y) to node {$1$} (Y);
  \draw [path] (1) to node {$\mu_X$} (X);
  \draw [path] (1) to node {$\delta_M$} (M);
  \draw [path] (1) to node {$\delta_Y$} (Y);
  \draw [cov] (X) to[out=170,in=190,looseness=5] node[left] {$\sigma^{2}_{X}$} (X);
  \draw [cov] (epsilon_M) to[out=70,in=110,looseness=5] node[above] {$\sigma^{2}_{\epsilon_{M}}$} (epsilon_M);
  \draw [cov] (epsilon_Y) to[out=70,in=110,looseness=5] node[above] {$\sigma^{2}_{\epsilon_{Y}}$} (epsilon_Y);
  \draw [cov] (1) to[out=-60,in=-120,looseness=7] node[below] {1} (1);
\end{tikzpicture}
```

### Data Generating Function

\begin{equation}
  X,
  M,
  Y
  \sim
  \mathrm{MVN}
    \left[
      \boldsymbol{\mu}
      \left(
        \boldsymbol{\theta}
      \right),
      \boldsymbol{\Sigma}
      \left(
        \boldsymbol{\theta}
      \right)
    \right]
\end{equation}

\begin{equation}
  \boldsymbol{\mu}
  \left(
    \boldsymbol{\theta}
  \right)
  = 
  \begin{bmatrix}
    \mu_X \\
    \delta_M + \alpha \mu_X \\
    \delta_Y + \tau^{\prime} \mu_X + \beta \delta_M + \alpha \beta \mu_X
  \end{bmatrix}
\end{equation}

\begin{equation}
  \boldsymbol{\Sigma}
  \left(
    \boldsymbol{\theta}
  \right)
  =
  \begin{bmatrix}
    \sigma^{2}_{X} & \alpha \sigma_{X}^{2} &  \alpha \beta \sigma_{X}^{2} + \tau^{\prime} \sigma_{X}^{2} \\
    \alpha \sigma_{X}^{2}  & \alpha^{2} \sigma_{X}^{2} + \sigma_{\varepsilon_{M}}^{2} & \alpha^{2} \beta \sigma_{X}^{2} + \alpha \tau^{\prime} \sigma_{X}^{2} + \beta \sigma_{\varepsilon_{M}}^{2} \\
    \alpha \beta \sigma_{X}^{2} + \tau^{\prime} \sigma_{X}^{2} & \alpha^{2} \beta \sigma_{X}^{2} + \alpha \tau^{\prime} \sigma_{X}^{2} + \beta \sigma_{\varepsilon_{M}}^{2} & \beta^{2} \alpha^{2} \sigma_{X}^{2} + \beta^{2} \sigma_{\varepsilon_{M}}^{2} + 2 \alpha \beta \tau^{\prime} \sigma_{X}^{2} + \tau^{\prime} \sigma_{X}^{2} + \sigma_{\varepsilon_{Y}}^{2}
  \end{bmatrix}
\end{equation}

```{r, echo = FALSE}
Variable <- c(
  "`muX`",
  "`deltaM`",
  "`deltaY`"
)
Description <- c(
  "Mean of $X$.",
  "Intercept of $M$.",
  "Intercept of $Y$."
)
Notation <- c(
  "$\\mu_X$",
  "$\\delta_M$",
  "$\\delta_Y$"
)
Value <- c(
  muX,
  deltaM,
  deltaY
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Mean of $X$ and Regression Intercepts"
)
```

```{r, echo = FALSE}
Variable <- c(
  "`alpha`",
  "`tauprime`",
  "`beta`",
  "`alphabeta`"
)
Description <- c(
  "Regression slope of path from $X$ to $M$.",
  "Regression slope of path from $X$ to $Y$.",
  "Regression slope of path from $M$ to $Y$.",
  "Indirect effect."
)
Notation <- c(
  "$\\alpha$",
  "$\\tau^{\\prime}$",
  "$\\beta$",
  "$\\alpha \\beta$"
)
Value <- c(
  alpha,
  tauprime,
  beta,
  alphabeta
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Regression Slopes"
)
```

```{r, echo = FALSE}
Variable <- c(
  "`sigma2X`",
  "`sigma2epsilonM`",
  "`sigma2epsilonY`"
)
Description <- c(
  "Variance of $X$.",
  "Error variance of $\\varepsilon_M$.",
  "Error variance of $\\varepsilon_Y$."
)
Notation <- c(
  "$\\sigma^2_X$",
  "$\\sigma^2_{\\varepsilon_{M}}$",
  "$\\sigma^2_{\\varepsilon_{Y}}$"
)
Value <- c(
  sigma2X,
  sigma2epsilonM,
  sigma2epsilonY
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Variance of $X$ and Error Variances"
)
```

```{r, echo = FALSE}
knitr::kable(
  x = mutheta,
  caption = "$\\boldsymbol{\\mu} \\left( \\boldsymbol{\\theta} \\right)$ (`mutheta`)"
)
knitr::kable(
  x = Sigmatheta,
  caption = "$\\boldsymbol{\\Sigma} \\left( \\boldsymbol{\\theta} \\right)$ (`Sigmatheta`)"
)
```

## Sample Data

```{r}
X <- as.data.frame(
  mvrnorm(
    n = n,
    Sigma = Sigmatheta,
    mu = mutheta
  )
)
str(X)
```

```{r, echo = FALSE}
psych::pairs.panels(
  X,
  method = "pearson", # correlation method
  hist.col = "#00AFBB",
  density = TRUE, # show density plots
  ellipses = TRUE # show correlation ellipses
)
```

### Sample Statistics

```{r, echo = FALSE}
muhat <- colMeans(X)
Sigmahat <- cov(X)
knitr::kable(
  x = muhat,
  caption = "Estimated Mean Vector $\\boldsymbol{\\hat{\\mu}}$ (`muhat`)"
)
knitr::kable(
  x = Sigmahat,
  caption = "Estimated Variance-Covariance Matrix $\\boldsymbol{\\hat{\\Sigma}}$ (`Sigmahat`)"
)
```

```{r}
model_01 <- lm(
  M ~ X,
  data = X
)
summary(model_01)
model_02 <- lm(
  Y ~ X + M,
  data = X
)
summary(model_02)
```

```{r, echo = FALSE}
muXhat <- mean(X$X)
sigma2Xhat <- var(X$X)
deltaMhat <- coef(model_01)[1]
alphahat <- coef(model_01)[2]
sigma2epsilonMhat <- (summary(model_01)$sigma)^2
deltaYhat <- coef(model_02)[1]
tauprimehat <- coef(model_02)[2]
betahat <- coef(model_02)[3]
sigma2epsilonYhat <- (summary(model_02)$sigma)^2
alphahatbetahat <- alphahat * betahat
thetahat <- alphahatbetahat
```

```{r, echo = FALSE}
Variable <- c(
  "`muXhat`",
  "`deltaMhat`",
  "`deltaYhat`"
)
Description <- c(
  "Estimated mean of $X$.",
  "Estimated intercept of $M$.",
  "Estimated intercept of $Y$."
)
Notation <- c(
  "$\\hat{\\mu_X}$",
  "$\\hat{\\delta_M}$",
  "$\\hat{\\delta_Y}$"
)
Value <- c(
  muXhat,
  deltaMhat,
  deltaYhat
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Sample Mean of $X$ and Regression Intercepts (Parameter Estimates)"
)
```

```{r, echo = FALSE}
Variable <- c(
  "`alphahat`",
  "`tauprimehat`",
  "`betahat`",
  "`alphahatbetahat`"
)
Description <- c(
  "Estimated regression slope of path from $X$ to $M$.",
  "Estimated regression slope of path from $X$ to $Y$.",
  "Estimated regression slope of path from $M$ to $Y$.",
  "Estimated indirect effect."
)
Notation <- c(
  "$\\hat{\\alpha}$",
  "$\\hat{\\tau}^{\\prime}$",
  "$\\hat{\\beta}$",
  "$\\hat{\\alpha}\\hat{\\beta}$"
)
Value <- c(
  alphahat,
  tauprimehat,
  betahat,
  alphahatbetahat
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Sample Regression Slopes (Parameter Estimates)"
)
```

```{r, echo = FALSE}
Variable <- c(
  "`sigma2Xhat`",
  "`sigma2epsilonMhat`",
  "`sigma2epsilonYhat`"
)
Description <- c(
  "Estimated variance of $X$.",
  "Estimated error variance of $\\varepsilon_M$.",
  "Estimated error variance of $\\varepsilon_Y$."
)
Notation <- c(
  "$\\hat{\\sigma}^2_X$",
  "$\\hat{\\sigma}^2_{\\varepsilon_{M}}$",
  "$\\hat{\\sigma}^2_{\\varepsilon_{Y}}$"
)
Value <- c(
  sigma2Xhat,
  sigma2epsilonMhat,
  sigma2epsilonYhat
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Variance of $X$ and Error Variances"
)
```

## Nonparametric Bootstrapping

### Bootstrap Samples

```{r}
s <- function(X) {
  model_01 <- lm(
    M ~ X,
    data = X
  )
  model_02 <- lm(
    Y ~ X + M,
    data = X
  )
  coef(model_01)[2] * coef(model_02)[3]
}
```

```{r}
X_star <- nb(
  data = X,
  B = B
)
```

### Bootstrap Sampling Distribution

```{r}
thetahat_star <- sapply(
  X = X_star,
  FUN = s
)
```

```{r, echo = FALSE}
hist(
  thetahat_star,
  main = expression(
    paste(
      "Histogram of ",
      hat(theta),
      "*"
    )
  ),
  xlab = expression(
    paste(
      hat(theta),
      "*"
    )
  )
)
qqnorm(thetahat_star)
qqline(thetahat_star)
```

### Estimated Bootstrap Standard Error

```{r, echo = FALSE}
mean_thetahat_star <- mean(thetahat_star)
var_thetahat_star <- var(thetahat_star)
sd_thetahat_star <- sqrt(var_thetahat_star)
```

The estimated bootstrap standard error
is given by

\begin{equation}
  \widehat{\mathrm{se}}_{\mathrm{B}}
  \left(
    \hat{\theta}
  \right)
  =
  \sqrt{
    \frac{1}{B - 1}
    \sum_{b = 1}^{B}
    \left[
      \hat{\theta}^{*} \left( b \right)
      -
      \hat{\theta}^{*} \left( \cdot \right)
    \right]^2
  }
  =
  `r sd_thetahat_star`
\end{equation}

\noindent where

\begin{equation}
  \hat{\theta}^{*}
  \left(
    \cdot
  \right)
  =
  \frac{1}{B}
  \sum_{b = 1}^{B}
  \hat{\theta}^{*} \left( b \right) 
  =
  `r mean_thetahat_star`
  .
\end{equation}

Note that
$\widehat{\mathrm{se}}_{\mathrm{B}} \left( \hat{\theta} \right)$
is the standard deviation of
$\boldsymbol{\hat{\theta}^{*}}$
and
$\hat{\theta}^{*} \left( \cdot \right)$
is the mean of
$\boldsymbol{\hat{\theta}^{*}}$ .

```{r, echo = FALSE}
Variable <- c(
  "`B`",
  "`mean_thetahat_star`",
  "`var_thetahat_star`",
  "`sd_thetahat_star`"
)
Description <- c(
  "Number of bootstrap samples.",
  "Mean of $B$ sample means.",
  "Variance of $B$ sample means.",
  "Standard deviation of $B$ sample means."
)
Notation <- c(
  "$B$",
  "$\\hat{\\theta}^{*} \\left( \\cdot \\right) = \\frac{1}{B} \\sum_{b = 1}^{B} \\hat{\\theta}^{*} \\left( b \\right)$",
  "$\\widehat{\\mathrm{Var}}_{\\mathrm{B}} \\left( \\hat{\\theta} \\right) = \\frac{1}{B - 1} \\sum_{b = 1}^{B} \\left[ \\hat{\\theta}^{*} \\left( b \\right) - \\hat{\\theta}^{*} \\left( \\cdot \\right) \\right]^2$",
  "$\\widehat{\\mathrm{se}}_{\\mathrm{B}} \\left( \\hat{\\theta} \\right) = \\sqrt{ \\frac{1}{B - 1} \\sum_{b = 1}^{B} \\left[ \\hat{\\theta}^{*} \\left( b \\right) - \\hat{\\theta}^{*} \\left( \\cdot \\right) \\right]^2 }$"
)
Value <- c(
  B,
  mean_thetahat_star,
  var_thetahat_star,
  sd_thetahat_star
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Notation,
    Value
  ),
  row.names = FALSE,
  caption = "Nonparametric Bootstrapping Results"
)
```

### Confidence Intervals

Confidence intervals can be constructed
around
$\hat{\theta}$ .
The functions
`pc()`,
`bc()`,
and
`bca()`
from
the `jeksterslabRboot` package
can be used to construct
confidence intervals
using the default alphas
of
0.001,
0.01,
and
0.05.
The confidence intervals can also be evaluated.
Since we know the population parameter theta
$\left(\theta = \mu = `r theta` \right)$,
we can check if our confidence intervals contain
the population parameter.

See documentation for
`pc()`,
`bc()`,
and
`bca()`
from
the `jeksterslabRboot` package
on how confidence intervals are constructed.

See documentation for
`zero_hit()`,
`theta_hit()`,
`len()`,
and
`shape()`
from
the `jeksterslabRboot` package
on how confidence intervals are evaluated.

```{r}
pc_out <- pc(
  thetahat_star = thetahat_star,
  thetahat = thetahat,
  eval = TRUE,
  theta = theta
)
bc_out <- bc(
  thetahat_star = thetahat_star,
  thetahat = thetahat,
  eval = TRUE,
  theta = theta
)
bca_out <- bca(
  thetahat_star = thetahat_star,
  thetahat = thetahat,
  data = X,
  fitFUN = s,
  eval = TRUE,
  theta = theta
)
```

```{r, echo = FALSE}
knitr::kable(
  x = as.data.frame(
    rbind(
      pc = pc_out,
      bc = bc_out,
      bca = bca_out
    )
  ),
  caption = "Confidence Intervals"
)
```

## Notes

## References

- Bollen, K. A., & Stine, R., (1990).
  Direct and indirect effects: Classical and bootstrap estimates of variability.
  *Sociological Methodology*, *20*, 115-40.
  https://doi.org/10.2307/271084
- Efron, B., & Tibshirani, R. J. (1993).
  *An introduction to the bootstrap*.
  New York, N.Y: Chapman & Hall.
- Shrout, P. E., & Bolger, N.  (2002).
  Mediation in experimental and nonexperimental studies: New procedures and recommendations.
  *Psychological Methods*, *7*, 422-445.
  https://doi.org/10.1037/1082-989X.7.4.422
- [Wikipedia: Bootstrapping (statistics)](https://en.wikipedia.org/wiki/Bootstrapping_(statistics))

