---
title: "Notes: Introduction to Parametric Bootstrapping"
author: "Ivan Jacob Agaloos Pesigan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes: Introduction to Parametric Bootstrapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(jeksterslabRboot)
```

## Random Variable

```{r, echo = FALSE}
set.seed(42)
n <- 100
mu <- 0
sigma2 <- 1
sigma <- sqrt(sigma2)
B <- 100
```

Let $X$ be a random variable
from a normal distribution with $\mu = `r mu`$
and $\sigma = `r sigma`$.

\begin{equation}
  X
  \sim
  \mathcal{N}
  \left(
    \mu = `r mu`,
    \sigma^2 = `r sigma2`
  \right)
\end{equation}

This data generating function is referred to as $F$.

## Realization of the Random Variable

Let $\mathbf{x}$ be a realization of the random variable $X$. 

```{r}
x <- rnorm(
  n = n,
  mean = mu,
  sd = sigma
)
```

```{r, echo = FALSE}
muhat <- mean(x)
se_muhat <- sd(x) / sqrt(n)
sigma2hat <- var(x)
sigmahat <- sd(x)
Variable <- c(
  "`n`",
  "`muhat`",
  "`se_muhat`",
  "`sigma2hat`",
  "`sigmahat`"
)
Description <- c(
  "Sample size ($n$)",
  "Estimated mean ($\\hat{\\mu}$)",
  "Standard error of the mean ($\\mathrm{se} \\left( \\hat{\\mu} \\right)$)",
  "Estimated variance ($\\hat{\\sigma}^2$)",
  "Estimated standard deviation ($\\hat{\\sigma}$)"
)
Value <- c(
  n,
  muhat,
  se_muhat,
  sigma2hat,
  sigmahat
)
knitr::kable(
  x = data.frame(
    Variable,
    Description,
    Value
  ),
  row.names = FALSE,
  caption = "Parameter Estimates"
)
```

```{r, echo = FALSE}
hist(x)
qqnorm(x)
qqline(x)
```

## Parametric Bootstrapping

$\mathbf{x}$ which is a realization of the random variable $X$
is used in bootstrapping.
This sample data is referred to as the empirical distribution
$\hat{F}$.

In parametric bootstrapping we make a parametric assumption
about the distribution of $\hat{F}$.
In the current example,
we assume a normal distribution $\hat{F}_{\mathcal{N} \left( \mu, \sigma^2 \right)}$.
Data is generated $B$ number of times (`r B` in the current example)
from the empirical distribution 
$\hat{F}_{\mathcal{N} \left( \mu, \sigma^2 \right)}$
using parameter estimates
($\hat{\mu} = `r muhat`$ and $\hat{\sigma}^2 = `r sigma2hat`$).

### $x^{*} = \left\{ x^{*1}, x^{*2}, x^{*3}, x^{*b}, \dots, x^{*B} \right\}$

$x^{*}$ is a set of $B$ bootstrap samples.

```{r}
x_star <- pb_univ(
  rFUN = rnorm,
  n = n,
  B = B,
  mean = muhat,
  sd = sigmahat
)
```

### $\hat{\theta}^{*} = \left\{ \hat{\theta}^{*} \left( 1 \right), \hat{\theta}^{*} \left( 2 \right), \hat{\theta}^{*} \left( 3 \right), \hat{\theta}^{*} \left( b \right), \dots, \hat{\theta}^{*} \left( B \right)\right\}$

The parameter is estimated for each of the $B$
bootstrap samples.
For example,
if we are interested in making inferences about the mean,
we estimate the mean $B$ times.
The estimated parameters form the bootstrap sampling distribution
$\hat{\theta}^{*}$.

```{r}
thetahat_star <- lapply(
  X = x_star,
  FUN = mean
)
thetahat_star <- as.vector(
  do.call(
    what = "rbind",
    args = thetahat_star
  )
)
```

```{r, echo = FALSE}
hist(thetahat_star)
qqnorm(thetahat_star)
qqline(thetahat_star)
```

### $\mathrm{se} \left( \hat{\theta}^{*} \right)$

```{r, echo = FALSE}
se_thetahat_star <- sd(thetahat_star)
mean_thetahat_star <- mean(thetahat_star)
```

The standard error of the bootstrap sampling distribution
$\hat{\theta}^{*}$
is given by

\begin{equation}
  \mathrm{se}
  \left(
    \hat{\theta}^{*}
  \right)
  =
  \sqrt{
    \frac{1}{B - 1}
    \sum_{b = 1}^{B}
    \left[
      \hat{\theta}^{*} \left( b \right)
      -
      \hat{\theta}^{*} \left( \cdot \right)
    \right]^2
  }
  =
  `r se_thetahat_star`
\end{equation}

where
the mean of the bootstrap sampling distribution
$\hat{\theta}^{*}$
is given by

\begin{equation}
  \hat{\theta}^{*}
  \left(
    \cdot
  \right)
  =
  \frac{1}{B}
  \sum_{b = 1}^{B}
  \hat{\theta}^{*} \left( b \right) 
  =
  `r mean_thetahat_star`
  .
\end{equation}

The standard error can be used to make inferences about $\hat{\theta}$.

### Confidence Intervals

Confidence intervals can be constructed
to make inferences about the stability of $\hat{\theta}$.

```{r}
wald <- wald(
  thetahat = muhat,
  se_thetahat = se_muhat,
  theta_null = 0,
  alpha = c(
    0.001,
    0.01,
    0.05
  ),
  distribution = "z"
)
pc <- pc(
  thetahat_star = thetahat_star,
  alpha = c(
    0.001,
    0.01,
    0.05
  ),
  wald = TRUE,
  thetahat = muhat,
  theta_null = 0,
  distribution = "z"
)
bc <- bc(
  thetahat_star = thetahat_star,
  thetahat = muhat,
  alpha = c(
    0.001,
    0.01,
    0.05
  ),
  wald = TRUE,
  theta_null = 0,
  distribution = "z"
)
bca <- bca(
  thetahat_star = thetahat_star,
  thetahat = muhat,
  data = x,
  fitFUN = mean,
  alpha = c(
    0.001,
    0.01,
    0.05
  ),
  wald = TRUE,
  theta_null = 0,
  distribution = "z"
)
```

```{r}
knitr::kable(
  x = as.data.frame(
    rbind(
      wald,
      pc,
      bc,
      bca
    )
  ),
  caption = "Confidence Intervals"
)
```
